---
title: 'How nhlscraper Scores Expected Goals'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How nhlscraper Scores Expected Goals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = '#>',
  fig.align = 'center',
  out.width = '92%',
  fig.width = 7,
  fig.height = 4.5
)

make_table <- function(x, caption, digits = 3) {
  knitr::kable(x, caption = caption, digits = digits)
}
```

## What xG Means Here

Expected goals, or xG, is the estimated probability that a shot becomes a goal.
In `nhlscraper`, `calculate_expected_goals()` adds one column, `xG`, to a
current-schema play-by-play table. Non-shot rows receive `NA`; shot-attempt rows
receive probabilities from bundled XGBoost models.

The package does **not** train models during package use. It ships frozen model
bundles that were trained outside the package and then stored with the runtime
preprocessing contract. That contract is just as important as the boosters:
numeric medians, categorical levels, dummy-column maps, and final feature order
all have to match training exactly.

## Basic Use

```{r basic-use, eval = FALSE}
pbp <- nhlscraper::gc_play_by_play(2023030417)
pbp <- nhlscraper::add_shift_times(
  play_by_play = pbp,
  shift_chart  = nhlscraper::shift_chart(2023030417)
)
pbp <- nhlscraper::add_deltas(pbp)
pbp <- nhlscraper::calculate_expected_goals(pbp)
```

`calculate_expected_goals()` can derive several missing context columns itself,
but the richest input is a play-by-play that already has shift timing and
event-to-event deltas. The function keeps the legacy `model` argument for
compatibility, but that argument is ignored.

## Six Shot Environments

The model system is not one giant all-purpose classifier. Each target-season
vintage contains six mutually exclusive models:

```{r partition-table, echo = FALSE}
partition_table <- data.frame(
  partition = c('sd', 'ev', 'pp', 'sh', 'en', 'ps'),
  name = c(
    'Standard 5v5',
    'Other even strength',
    'Power play',
    'Short-handed',
    'Empty net',
    'Penalty shot / shootout'
  ),
  rows_sent_there = c(
    'Regulation 5v5 shots with both goalies in net, plus safe fallbacks.',
    'Remaining even-strength shots such as 4v4 and 3v3.',
    'Shots where the shooting team has a skater advantage.',
    'Shots where the shooting team has fewer skaters.',
    'Shots at an empty opposing net.',
    'Penalty-shot and shootout-style one-on-one attempts.'
  ),
  stringsAsFactors = FALSE
)
make_table(
  partition_table,
  caption = 'Shot partitions used by calculate_expected_goals().'
)
```

The order matters. A shootout attempt is handled before empty-net or manpower
rules. Empty-net shots are pulled out before normal strength partitions. Standard
5v5 is separated from other even-strength play because its sample is large and
its scoring environment is cleaner.

## Runtime Routing

Each shot is routed by game season and game state:

```{r routing-plot, echo = FALSE, fig.width = 7.5, fig.height = 4.8, fig.cap = 'Runtime routing from play-by-play row to xG value.'}
steps <- c(
  'Shot row',
  'Target season',
  'Partition',
  'Frozen recipe',
  'XGBoost booster',
  'xG'
)
box_x <- c(1, 2, 3, 3, 2, 1)
box_y <- c(2, 2, 2, 1, 1, 1)
box_w <- 0.78
box_h <- 0.36
draw_box <- function(x, y, label, number) {
  wrapped_label <- paste(strwrap(paste0(number, '. ', label), width = 16), collapse = '\n')
  graphics::rect(
    xleft   = x - box_w / 2,
    ybottom = y - box_h / 2,
    xright  = x + box_w / 2,
    ytop    = y + box_h / 2,
    col     = '#fefae0',
    border  = '#bc6c25',
    lwd     = 1.5
  )
  graphics::text(
    x      = x,
    y      = y,
    labels = wrapped_label,
    cex    = 0.84,
    col    = '#1f2933'
  )
}
draw_arrow <- function(i, j) {
  x0 <- box_x[i]
  y0 <- box_y[i]
  x1 <- box_x[j]
  y1 <- box_y[j]
  graphics::arrows(
    x0     = x0 + sign(x1 - x0) * box_w / 2,
    y0     = y0 + sign(y1 - y0) * box_h / 2,
    x1     = x1 - sign(x1 - x0) * box_w / 2,
    y1     = y1 - sign(y1 - y0) * box_h / 2,
    length = 0.08,
    lwd    = 1.3,
    col    = '#495057'
  )
}
graphics::plot(
  NA_real_,
  NA_real_,
  type = 'n',
  axes = FALSE,
  xlab = '',
  ylab = '',
  xlim = c(0.45, 3.55),
  ylim = c(0.55, 2.45)
)
for (i in seq_len(length(steps) - 1L)) {
  draw_arrow(i, i + 1L)
}
for (i in seq_along(steps)) {
  draw_box(box_x[i], box_y[i], steps[i], i)
}
```

Historical games use the target-season vintage when one exists. Seasons before
the bundled range use the earliest available vintage. Seasons beyond the bundled
range use the latest deployment vintage. That behavior keeps scoring possible
for old and future rows while preserving rolling-model logic where exact
vintages exist.

## Feature Families

The feature set is intentionally broader than "distance plus angle." The model
frame includes information about where the shot came from, what happened just
before it, who took it, who was in net, and what state the game was in.

```{r feature-table, echo = FALSE}
feature_table <- data.frame(
  family = c(
    'Shot geometry',
    'Shot location bins',
    'Previous-event movement',
    'Rush and rebound context',
    'Game state',
    'Strength state',
    'Shooter and goalie biometrics',
    'Shift timing',
    'Shootout counters'
  ),
  examples = c(
    'x/y, normalized x/y, distance, angle',
    'slot, net-front, point, flank, perimeter indicators',
    'delta seconds, delta x/y, delta distance, delta angle',
    'isRush, isRebound, createdRebound, previous event type',
    'score differential, cumulative shots/Fenwick/Corsi',
    'skater counts, manpower differential, empty-net flags',
    'height, weight, handedness where available',
    'seconds elapsed/remaining in shift for on-ice players',
    'attempt order for one-on-one partitions'
  ),
  stringsAsFactors = FALSE
)
make_table(
  feature_table,
  caption = 'Feature families used by the bundled xG models.'
)
```

Not every partition uses every feature in the same way, and not every row has
every upstream field. The preprocessing bundle is responsible for converting the
available public schema into the exact numeric matrix the booster expects.

## Training Windows

Each completed target-season vintage is trained only on earlier seasons. For a
target season, the training window is the three immediately previous seasons.
That keeps the evaluation leak-free: the model never trains on the season it is
being evaluated against.

```{r training-table, echo = FALSE}
training_table <- data.frame(
  target_vintage = c('2013-14', '2018-19', '2023-24', '2026-27 deployment'),
  training_window = c(
    'Earliest bundled historical window',
    '2015-16, 2016-17, 2017-18',
    '2020-21, 2021-22, 2022-23',
    '2023-24, 2024-25, 2025-26'
  ),
  note = c(
    'Uses the earliest supported vintage behavior.',
    'Example completed rolling vintage.',
    'Example modern completed rolling vintage.',
    'Latest deployment model used for future/default scoring.'
  ),
  stringsAsFactors = FALSE
)
make_table(
  training_table,
  caption = 'Examples of rolling training windows.'
)
```

## Deployment Vintage Size

The latest deployment vintage is trained on a large three-season sample, but the
six partitions differ dramatically in size and base goal rate.

```{r deployment-training-table, echo = FALSE}
deployment_training <- data.frame(
  partition = c('sd', 'ev', 'pp', 'sh', 'en', 'ps'),
  train_seasons = rep('2023-24, 2024-25, 2025-26', 6),
  rows = c(283688, 7654, 59254, 8186, 2891, 2027),
  goals = c(16881, 813, 5678, 595, 1596, 645),
  goal_rate = c(0.0595, 0.1062, 0.0958, 0.0727, 0.5521, 0.3182),
  stringsAsFactors = FALSE
)
make_table(
  deployment_training,
  caption = 'Training volume for the shipped 2026-27 deployment vintage.',
  digits = 4
)
```

This is why the partitions exist. Empty-net attempts and penalty shots are not
rare versions of ordinary five-on-five shots; they are different scoring
problems with different base rates.

## Completed-Season Evaluation

Completed-season evaluation currently covers target seasons from `2013-14`
through `2025-26`.

```{r season-results, echo = FALSE}
season_results <- data.frame(
  season = c(
    '2013-14', '2014-15', '2015-16', '2016-17', '2017-18',
    '2018-19', '2019-20', '2020-21', '2021-22', '2022-23',
    '2023-24', '2024-25', '2025-26'
  ),
  rows = c(
    112051, 110922, 110263, 111708, 120543, 118438, 105028,
    79111, 122341, 122701, 123126, 120445, 120129
  ),
  goal_rate = c(
    0.0670, 0.0665, 0.0660, 0.0660, 0.0679, 0.0697, 0.0701,
    0.0712, 0.0730, 0.0736, 0.0712, 0.0714, 0.0736
  ),
  xg_rate = c(
    0.0665, 0.0664, 0.0669, 0.0666, 0.0664, 0.0674, 0.0694,
    0.0690, 0.0730, 0.0764, 0.0720, 0.0693, 0.0761
  ),
  roc_auc = c(
    0.7868, 0.7807, 0.7814, 0.7767, 0.7793, 0.7790, 0.7791,
    0.7843, 0.7756, 0.7685, 0.7737, 0.7812, 0.7945
  ),
  calibration_ratio = c(
    1.0065, 1.0011, 0.9876, 0.9918, 1.0224, 1.0328, 1.0093,
    1.0332, 1.0012, 0.9626, 0.9899, 1.0309, 0.9669
  ),
  stringsAsFactors = FALSE
)
make_table(
  season_results,
  caption = 'Completed-season xG evaluation by target season.',
  digits = 4
)
```

```{r evaluation-plot, echo = FALSE, fig.cap = 'Observed goal rate and xG rate by completed target season.'}
old_par <- graphics::par(no.readonly = TRUE)
graphics::par(mar = c(7, 4, 3, 1))
graphics::plot(
  seq_len(nrow(season_results)),
  season_results[['goal_rate']],
  type = 'b',
  pch = 19,
  lwd = 2,
  col = '#1d3557',
  xaxt = 'n',
  ylim = range(c(season_results[['goal_rate']], season_results[['xg_rate']])),
  xlab = '',
  ylab = 'Rate'
)
graphics::lines(
  seq_len(nrow(season_results)),
  season_results[['xg_rate']],
  type = 'b',
  pch = 17,
  lwd = 2,
  col = '#e63946'
)
graphics::axis(
  side = 1,
  at = seq_len(nrow(season_results)),
  labels = season_results[['season']],
  las = 2,
  cex.axis = 0.75
)
graphics::mtext('Target Season', side = 1, line = 5)
graphics::legend(
  'topleft',
  legend = c('Observed goal rate', 'Average xG'),
  col = c('#1d3557', '#e63946'),
  pch = c(19, 17),
  lwd = 2,
  bty = 'n'
)
graphics::par(old_par)
```

Across completed seasons, ROC AUC ranges from `0.7685` to `0.7945`, and the
calibration ratio ranges from `0.9626` to `1.0332`. Those values are not a
promise that every game-level sum will be exact. They are a check that, across
large seasonal samples, the model stays close to observed scoring rates while
preserving useful ranking power.

## Caveats

Use xG as an estimate of chance quality, not as a perfect replay of intent. The
model sees public event and tracking-derived context. It does not see every
screen, pre-shot pass, goalie sightline, defensive stick, shooter injury, or
tactical instruction. The best use is comparative:

- Which team created more dangerous attempts?
- Which period changed the game?
- Which players produced the best looks?
- Did a club win by shot volume, shot quality, or finishing?

## Key Takeaway

`calculate_expected_goals()` is intentionally simple at the user level and more
careful under the hood. Give it a current-schema play-by-play, and it routes
each shot through a rolling season vintage, a game-state partition, a frozen
preprocessing recipe, and a bundled XGBoost booster. The returned `xG` column is
therefore easy to use, but it is not a black box stapled onto raw NHL data.
