---
title: "How nhlscraper's Expected Goals Model Works"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How nhlscraper's Expected Goals Model Works}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

make_table <- function(x, caption, digits = 3) {
  knitr::kable(x, caption = caption, digits = digits)
}
```

## Overview

Expected goals, or xG, estimates the probability that a shot becomes a goal. `nhlscraper` now ships rolling XGBoost xG models trained outside the package in the companion model repository and bundled for package runtime scoring. The package does not train models locally. It prepares public play-by-play rows, routes each shot to the appropriate season vintage and game-state partition, bakes the frozen preprocessing matrix, and asks the bundled XGBoost booster for a probability.

The shipped bundle contains six models for every target season from `2013-14` through the `2026-27` deployment vintage. Historical games use their matching target-season vintage when available. Seasons before the bundled range use the earliest vintage, and seasons after the bundled range use the latest deployment vintage.

## Six Situations

Each vintage is split into six mutually exclusive shot environments:

```{r partition-table}
partition_table <- data.frame(
  partition = c("sd", "ev", "pp", "sh", "en", "ps"),
  meaning = c(
    "Regulation 5v5 without empty nets, plus fallback rows",
    "Other even-strength states outside standard 5v5",
    "Shooting team has a skater advantage",
    "Shooting team is short-handed",
    "Opponent net is empty",
    "Penalty-shot and shootout-style situations"
  ),
  stringsAsFactors = FALSE
)
make_table(
  partition_table,
  caption = "The six shot partitions used by nhlscraper's xG models."
)
```

The scorer applies these rules in order:

1. Penalty-shot and shootout-style states (`1010` and `0101`) go to `ps`.
2. Empty-net-against shots go to `en`.
3. Standard 5v5 non-empty-net shots go to `sd`.
4. Remaining even-strength shots go to `ev`.
5. Skater-advantage shots go to `pp`.
6. Skater-disadvantage shots go to `sh`.

That split matters because a 5v5 point shot, a 4v4 rush, a power-play one-timer, an empty-net attempt, and a shootout attempt are different statistical problems.

## Training Design

Every target-season model is leak-free: it is trained only on the three immediately previous seasons. For example, the `2026-27` deployment vintage was trained on `2023-24`, `2024-25`, and `2025-26`. Hyperparameters are selected upstream with capped rolling Bayesian optimization, then the chosen parameters are refit on the full three-season window for that vintage and partition.

```{r deployment-training-table}
deployment_training <- data.frame(
  partition = c("sd", "ev", "pp", "sh", "en", "ps"),
  train_seasons = rep("2023-24, 2024-25, 2025-26", 6),
  rows = c(283688, 7654, 59254, 8186, 2891, 2027),
  goals = c(16881, 813, 5678, 595, 1596, 645),
  goal_rate = c(0.0595, 0.1062, 0.0958, 0.0727, 0.5521, 0.3182),
  stringsAsFactors = FALSE
)
make_table(
  deployment_training,
  caption = "Training volume for the shipped 2026-27 deployment vintage.",
  digits = 4
)
```

## Runtime Features

The runtime scorer starts from the current public play-by-play schema. When needed, it adds the same context used by the training pipeline:

```r
pbp <- nhlscraper::gc_pbps(season) |>
  nhlscraper::add_shift_times(nhlscraper::shift_charts(season)) |>
  nhlscraper::add_deltas() |>
  nhlscraper::add_shooter_biometrics() |>
  nhlscraper::add_goalie_biometrics()
```

The XGBoost models use more than shot location. They include normalized shot geometry, slot and net-front flags, event-to-event movement, rebound and rush flags, previous-event descriptors, score state, shot/Fenwick/Corsi context, skater counts, strength state, shooter and goalie biometrics, shift-timing summaries, and shootout counters for `ps`.

The package stores the preprocessing contract alongside the boosters: numeric medians, categorical levels, dummy-column maps, and feature order. That matters because XGBoost requires the runtime matrix to match training exactly.

## Runtime Path

For each shot, `calculate_expected_goals()` follows this path:

1. Validate that the play-by-play uses the current public schema.
2. Add missing context columns where the package can derive them.
3. Build the model frame for shot events.
4. Select the target-season vintage from `gameId`.
5. Partition each shot into `sd`, `ev`, `pp`, `sh`, `en`, or `ps`.
6. Bake the frozen preprocessing matrix for that vintage and partition.
7. Score the row with the bundled XGBoost booster and return `xG`.

The legacy `model` argument remains accepted for compatibility, but it is ignored.

## Evaluation

Historical target-season evaluation covers `2013-14` through `2025-26`. Across those completed seasons, ROC AUC ranges from `0.7685` to `0.7945`, and calibration ratio ranges from `0.9626` to `1.0332`.

```{r season-table}
season_results <- data.frame(
  season = c(
    "2013-14", "2014-15", "2015-16", "2016-17", "2017-18",
    "2018-19", "2019-20", "2020-21", "2021-22", "2022-23",
    "2023-24", "2024-25", "2025-26"
  ),
  rows = c(
    112051, 110922, 110263, 111708, 120543, 118438, 105028,
    79111, 122341, 122701, 123126, 120445, 120129
  ),
  goal_rate = c(
    0.0670, 0.0665, 0.0660, 0.0660, 0.0679, 0.0697, 0.0701,
    0.0712, 0.0730, 0.0736, 0.0712, 0.0714, 0.0736
  ),
  xg_rate = c(
    0.0665, 0.0664, 0.0669, 0.0666, 0.0664, 0.0674, 0.0694,
    0.0690, 0.0730, 0.0764, 0.0720, 0.0693, 0.0761
  ),
  log_loss = c(
    0.2084, 0.2095, 0.2070, 0.2091, 0.2124, 0.2165, 0.2183,
    0.2184, 0.2235, 0.2276, 0.2221, 0.2192, 0.2197
  ),
  roc_auc = c(
    0.7868, 0.7807, 0.7814, 0.7767, 0.7793, 0.7790, 0.7791,
    0.7843, 0.7756, 0.7685, 0.7737, 0.7812, 0.7945
  ),
  calibration_ratio = c(
    1.0065, 1.0011, 0.9876, 0.9918, 1.0224, 1.0328, 1.0093,
    1.0332, 1.0012, 0.9626, 0.9899, 1.0309, 0.9669
  )
)
make_table(
  season_results,
  caption = "Completed-season xG evaluation by target season.",
  digits = 4
)
```

The latest completed target season is `2025-26`. Its partition-level results show the expected pattern: standard 5v5 carries the largest and most stable sample, special-teams models remain useful, empty-net attempts behave like their own scoring environment, and shootout/penalty-shot attempts are noisier.

```{r latest-partition-table}
latest_partition_results <- data.frame(
  partition = c("sd", "ev", "pp", "sh", "en", "ps"),
  rows = c(92973, 2719, 19915, 2632, 1055, 835),
  goal_rate = c(0.0610, 0.0982, 0.0951, 0.0707, 0.5185, 0.3234),
  xg_rate = c(0.0615, 0.1051, 0.1032, 0.0786, 0.5662, 0.3323),
  log_loss = c(0.1935, 0.3036, 0.2962, 0.1949, 0.6035, 0.6301),
  roc_auc = c(0.8004, 0.7112, 0.6748, 0.8589, 0.7532, 0.5439),
  calibration_ratio = c(0.9925, 0.9339, 0.9211, 0.8988, 0.9158, 0.9731)
)
make_table(
  latest_partition_results,
  caption = "Latest completed-season (`2025-26`) results by partition.",
  digits = 4
)
```

## Practical Takeaways

`nhlscraper` no longer exposes xG as a menu of model versions. The package ships one coherent xG system made of rolling season vintages and six game-state partitions. For users, the interface stays simple: call `calculate_expected_goals()` on current-schema play-by-play data and read the returned `xG` column. Internally, the scorer now uses the same training contract as the upstream XGBoost models.
